Detection and partial characterisation of stellar multiplicity with Gaia

### 2018-08-23 article/ms.tex
trigger travis
minted etc

\documentclass[twocolumn]{aastex61}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{color}
\usepackage{comment}

@@

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\texttt{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\rp}{\textsl{rp}}
\newcommand{\bp}{\textsl{bp}}

\newcommand{\NumberOfStellarMultiples}{XXX}
\newcommand{\NumberOfStellarSingles}{XXX}

@@

@@

For a single star, or a star without a significant mass companion, the source radial
velocity jitter represents the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that colour, apparent magnitude, and absolute magnitude.
Stars with bluer colours have fewer absorption lines with deeper wings, and fainter stars
have on average lower signal-to-noise (S/N) ratios, both of which result in noisier radial
velocity measurements. The absolute magnitude is similarly important, as giant stars have
narrow absorption lines than main-sequence stars of the same temperature and colour. For stellar multiples, the source radial velocity jitter will be the quadrature sum of the expected jitter for the primary (if it were a single star) and a contribution based on the orbital parameters  and the observation epochs.

% todo: lack of RV measurements?

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) may have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a detectable jitter in the astrometric 
position. There are many other sources of astrometric jitter, but here we show 
that the principle effect of astrometric excess noise is due to stellar 
companions. Note that due to the astrometry `degree of freedom bug' in the 
processing of the second \Gaia\ data release \citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, 
that resulted in 80\% of sources having zero astrometric excess noise,
we instead make use of the astrometric unit weight error $u$
\begin{eqnarray}
	u = \sqrt{\frac{\chi^2_{al}}{N_{obs,al} - 5}} \label{eq:astrometric_unit_weight_error}
\end{eqnarray}
\noindent{}where $\chi^2_{al}$ is the astrometric $\chi^2$ value
in the along-scan direction (column \texttt{astrometric\_chi2\_al}) and 
$N_{obs,al}$ is the number of astrometric observations in the along-scan
direction (column \texttt{astrometric\_n\_obs\_al}). The values of this
quantity can be interpreted similarly to how the astrometric excess noise
would be interpreted (e.g., large values are less consistent with a
single star model), although like the radial velocity and astrometric
jitter, it must be considered in context with stars of similar colour,
apparent magnitude, and preferably, absolute magnitude.

% todo: what ADQL queries did we do?
% what will our subset be?

@@

We assume that for sources with similar \bp\ - \rp\ colours, apparent \rp\ magnitudes, and absolute \rp\ magnitudes, the jitter distribution (in radial velocity or astrometry) is a mixture of two populations:

\begin{enumerate}
\item \emph{single-star systems}, where the jitter represents the minimum noise that 
	   \Gaia\ can measure for a star of that
      \bp\ - \rp\ colour, apparent \rp\ magnitude, and absolute \rp\ magnitude;

\item \emph{higher-order star systems}, where the jitter is
      significantly above that for single-star systems of a similar colour, 
      apparent magnitude, and absolute magnitude.
\end{enumerate}

We also assume that the mean single-star jitter will change
depending on the \bp\ - \rp\ colour, the apparent magnitude, and the absolute magnitude.
For example, the mean radial velocity jitter of single stars with $\teff \approx 6000\,\textrm{K}$
will be higher than the jitter of single stars with $\teff \approx 5000\,\textrm{K}$, 
with all else being (nearly) equal. We similarly assume that the mean 
jitter of stellar \emph{multiples} will depend on the source colour and its position
in the Hertzsprung-Russell diagram. For these reasons, in order to evaluate whether a
source is more likely to be a single star or a stellar multiple, we must consider the 
jitter in context with other sources that have similar: \bp\ - \rp\
colour, apparent \rp\ magnitude, and absolute \rp\ magnitude. In principle if we knew
the distribution stellar multiple properties (or assumed one) then we could model the
observability of binary (and higher order) systems across the Hertszprung-Russell
diagram, but this is beyond the scope of this work.

% ADDITIONAL ASSUMPTIONS.
\todo{additional assumptions}

Given these assumptions, we chose to adopt a non-parametric model for stellar multiplicity,
primarily because we do not know the distribution of properties of stellar multiples.
Our approach is illustrated in Figure \ref{fig:npm_schematic} and can be qualitatively 
described by the following steps.\\

\noindent{}For each \Gaia\ source:
\begin{enumerate}

	\item Select $N$ sources that have a similar \bp\ - \rp\ colour, apparent \rp\
		  magnitude, and absolute \rp\ magnitude.

	\item Fit the distribution of radial velocity jitter for sources selected in
		  Step 1 using a mixture of a normal and a log-normal distribution.

	\item Evaluate whether the source of interest is more consistent with being
		  a single star component (drawn from the normal distribution) or a
		  stellar multiple (drawn from the log-normal distribution).

	\item If the source is more likely to be a stellar multiple, use the estimated
		  properties of single stars with similar properties to estimate the
		  excess radial velocity jitter due to the stellar companion(s).

\end{enumerate}

@@

\appendix
\section{Reproducibility}
This project was developed in a \texttt{git} repository that is publicly
accessible at \url{https://github.com/andycasey/velociraptor}. The repository
includes notebooks that detail our work, \LaTeX\ to compile this manuscript, 
and scripts to reproduce our analysis in its entirety. Reproducing this work in 
full will require at least \todo{X}\,Gb of disk space and \todo{Y}\,compute hours.
To reproduce this work (including data retrieval, analysis, production of figures,
and manuscript compilation) use the \todo{following commands on a modern terminal}:

\section{Relevant Astronomical Data Query Language (ADQL) queries}
\todo{Include relevant ADQL queries}

@@

\software{
	\package{Astropy} \citep{astropy},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub},
    \package{TensorFlow} \citep{tensorflow}
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}

>>>>

\documentclass[twocolumn]{aastex61}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{color}
\usepackage{comment}
\usepackage{minted}

@@

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\package}[1]{\texttt{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\ESA}{\acronym{ESA}}
\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\rp}{\textsl{rp}}
\newcommand{\bp}{\textsl{bp}}

\newcommand{\NumberOfStellarMultiples}{XXX}
\newcommand{\NumberOfStellarSingles}{XXX}
\newcommand{\NumberOfSourcesInSubset}{8,911,018} % check

@@

%%% This file is generated by the Makefile.

\newcommand{\vcpath}{vc.tex}

\IfFileExists{\vcpath}{\input{\vcpath}}{
	\newcommand{\giturl}{UNKNOWN}
	\newcommand{\gitslug}{UNKNOWN}
	\newcommand{\githash}{UNKNOWN}
	\newcommand{\gitdate}{UNKNOWN}
	\newcommand{\gitauthor}{UNKNOWN}
}

@@

For a single star, or a star without a significant mass companion, the source radial
velocity jitter represents the minimum uncertainty with which \Gaia\ can measure 
radial velocity for a star of that colour, apparent magnitude, and absolute magnitude.
Stars with bluer colours have fewer absorption lines with deeper wings, and fainter stars
have on average lower signal-to-noise (S/N) ratios, both of which result in noisier radial
velocity measurements. The absolute magnitude is similarly important, as giant stars have
narrow absorption lines than main-sequence stars of the same temperature and colour. For stellar multiples, the source radial velocity jitter will be the quadrature sum of the expected jitter for the primary (if it were a single star) and a contribution from the radial velocity semi-amplitude of the orbit.

Some sources in \Gaia\ do not have reported radial velocities even though they are
apparently bright enough and within a suitable \bp\ - \rp\ colour range.  The likely
explanation for these situations is that the \Gaia\  team identified
the source to be a double-lined spectroscopic binary, and there is no utility in
reporting a point estimate of the \emph{source} radial velocity and associated error.
Other explanations are also possible, which implies that we cannot use the \emph{lack}
of radial velocity as a robust identifier of stellar multiplicity. \todo{In the following
Sections}
% Show it ehre?
we demonstrate that for sources that do not have reported radial velocities -- and 
sources of similar apparent magnitude and colour \emph{do} have radial velocities --
their astrometric jitter is usually sufficient to distinguish single stars from stellar
multiples.

% todo:  Figure showing distribution of astrometric unit weight error.

@@

Many higher-order star systems are unresolved point sources in \Gaia, but the
stellar companion(s) may have sufficient mass to measurably affect the tangential 
velocity of the primary and produce a detectable jitter in the astrometric 
position. 
%There are many other sources of astrometric jitter, but here we show 
%that the principle effect of astrometric excess noise is due to stellar 
%companions.
We make use of the astrometric unit weight error $u$ to measure astrometric
jitter
\begin{eqnarray}
	u = \sqrt{\frac{\chi^2_{al}}{N_{obs,al} - 5}} \label{eq:astrometric_unit_weight_error}
\end{eqnarray}
\noindent{}where $\chi^2_{al}$ is the astrometric $\chi^2$ value
in the along-scan direction (column \texttt{astrometric\_chi2\_al}) and 
$N_{obs,al}$ is the number of astrometric observations in the along-scan
direction (column \texttt{astrometric\_n\_obs\_al}). 
The astrometric unit weight error is unaffected by the `degree of freedom bug'
that occurred during the processing of the the second \Gaia\ data release
\citep[e.g., see Appendix A1 of ][]{Lindegren:2018}, which resulted in 80\% of
sources having zero astrometric excess noise. 
The astrometric unit weight error can be interpreted similarly to how the
astrometric excess noise would be interpreted (e.g., large values are less
consistent with a single star model), but using the astrometric unit weight
error allows us to distinguish astrometric binaries in a much larger sample.
And like the radial velocity jitter, the unit weight astrometric error of a
single source must be considered in context with stars of similar colour 
and apparent magnitude (but not necessarily absolute magnitude).

% todo: what ADQL queries did we do?
% what will our subset be?
\subsection{\texttt{ADQL} query}

We restrict our work to the brightest $\sim{}10^7$ sources in \Gaia\ DR2.
About 70\% of sources in this subset have reported radial velocities,
allowing us to identify spectroscopic binaries, and this subset provides
a good sample of sources to test the precision to which we can identify
astrometric binaries. Specifically, we executed the following Astronomical Data Query
Language \citep[\texttt{ADQL};][]{someone} query on \ESA\ \Gaia\ archive
\citep{esa_gaia_archive}:

\begin{minted}[style=friendly]{postgresql}
SELECT *
  FROM gaiadr2.gaia_source
 WHERE radial_velocity IS NOT NULL
    OR phot_rp_mean_mag <= 13
\end{minted}
\noindent{}This query returned \NumberOfSourcesInSubset\ sources.

@@

We assume that for a sample of sources with similar \bp\ - \rp\ colours
and apparent \rp\ magnitudes -- and absolute \rp\ magnitudes, if radial velocities
are being considered -- the distribution of jitter is a mixture of two components:

\begin{enumerate}
\item \emph{single-star systems}, where the jitter represents the minimum noise that 
	   \Gaia\ can measure for a source with those properties (e.g., a source with
	   that \bp\ - \rp\ colour, apparent \rp\ magnitude, and perhaps absolute \rp\
	   magnitude);

\item \emph{higher-order star systems}, where the jitter is significantly higher 
	  than what is measured for single-star systems of similar properties.
\end{enumerate}

We further assume that the jitter for a single star will change depending on the
\bp\ - \rp\ colour, the apparent magnitude, and the absolute magnitude. For example, 
the mean radial velocity jitter of single stars with $\teff \approx 6000\,\textrm{K}$
will be higher than the jitter of single stars with $\teff \approx 5000\,\textrm{K}$, 
with all else being (nearly) equal. Similarly, the minimum radial velocity jitter
for giant stars is expected to be lower on average than for main-sequence stars of
the same colour, as giant stars have narrower absorption lines than main-sequence
stars of the same colour, allowing for a more precise measurement of the radial
velocity. However, this effect will not affect the astrometric jitter: the minimum
astrometric jitter for a main-sequence star and a giant star of the same temperature
is likely to be indistinguishable. For these reasons, in order to evaluate whether 
a source is more likely to be a single star or a stellar multiple, we must consider 
the jitter in context with other sources that have a similar \bp\ - \rp\ colour, 
apparent \rp\ magnitude, and in the case of radial velocity jitter, the absolute
\rp\ magnitude.\footnote{The magnitude in any band would be sufficient. Throughout
this work we chose the \rp\ band because its wavelength range most closely overlaps 
with that of the Radial Velocity Spectrograph on board \Gaia.} However, we cannot reliably describe a quantitative
relationship for \emph{how} exactly the astrometric and radial velocity jitter is
expected to vary for a single star given some source properties. For these reasons
we chose to adopt a flexible non-parametric model for stellar multiplicity, primarily
conditioned on the assumption that if we select enough sources with \emph{similar}
properties then within that subset of similar stars there exists \emph{some} single stars 
and \emph{some} stellar multiples.

We illustrate our approach in Figure \ref{fig:npm_schematic}, which can be qualitatively
described by the following steps. When considering the radial velocity jitter for a
\Gaia\ source of interest, we select $N$ sources that have a similar \bp\ - \rp\ colour,
apparent \rp\ magnitude, and absolute \rp\ magnitude. We then fit the distribution of
radial velocity jitter $\RVJitter$ as a mixture of two components: a normal distribution
to represent single stars, and a log-normal distribution to represent stellar multiples.
We perform these steps for $\approx10^4$ sources. While this process approximately maps
how the radial velocity jitter of single (and multiple) stars varies across the Hertszprung-Russell
diagram, each source of interest is considered independent from all others. For example,
for three sources that are all indistinguishable from each other in \bp\ - \rp\ colour, 
apparent \rp\ magnitude, and absolute \rp\ magnitude, the procedure we described above
treats the fitting of these three mixture models as entirely independent of each other,
even though we suspect that the parameters of the three mixture models ought to be 
identical. In other words: the procedure as described does not enforce any smoothness.
Given enough data this is not a problem for densely-populated regions of the 
Hertszprung-Russell diagram as the optimized results \emph{will} be (statistically)
indistinguishable for similar sources. In practice this poses a problem for poorly-sampled
regions of the Hertzsprung-Russell diagram, and raises subjective questions on what
constitutes \emph{the most} similar type of source.

\todo{In the third step we apply a Gaussian process.}

@@

\appendix
\section{Reproducibility}
This project was developed in a \texttt{git} repository that is publicly
accessible at \giturl. The repository includes notebooks that detail our work,  
\LaTeX\ to compile this manuscript, and scripts to reproduce the results presented in
this manuscript. The results presented here can be reproduced in full (including data
retrieval, analysis, creation of figures, and manuscript compilation) using these
commands on a modern terminal:
% this is not python, but latex fails to evaluate \githash if I set bash environment.
\begin{minted}[
style=friendly,
escapeinside=||
]{python} 
git clone https://github.com/andycasey/velociraptor.git velociraptor
cd velociraptor
git checkout |\githash|
./reproduce
\end{minted}

Reproducing these results will require at least \todo{X}\,Gb of free disk space 
and \todo{Y}\,hours of compute time. The settings in \texttt{model.yaml} can be 
adjusted to reduce the sample size of the data and shorten the compute time, but
varying these settings does not guarantee exact replication of the results shown
here.

@@

% astroquery
% minted
\software{
	\package{Astropy} \citep{astropy:v1,astropy:v2},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub},
    \package{TensorFlow} \citep{tensorflow}
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}
